MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-03-25 MAESTROeX_I/algorithm.tex
simplify alg description

To initialize the simulation we use the same procedure described in Paper III.
At the beginning of each simulation, we define $(\Ub,\rho X_k,\rho h)$.
We set initial values for $\Ub, \rho X_k$, and $\rho h$ and perform a sequence of projections 
(to ensure the velocity field satisfies the divergence constraint) 
followed by a small number of steps of the temporal advancement scheme to iteratively 
find initial values for $\pi^{n-\myhalf}$ and $S^0$ and $S^1$ for use in the first time step.
The new temporal integration strategy in MAESTROeX contains the following steps:
\MarginPar{mention cutoff densities}

\begin{description}

@@

\item Update the enthalpy using a discretized version of equation
%
\begin{equation}
\frac{\partial(\rho h)}{\partial t} = -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc,
\end{equation}
%
again omitting the reaction and heating terms
since we already accounted for
them in {\bf React State}.  This equation takes the form:
\begin{equation}
\frac{\partial (\rho h)}{\partial t}  = - \nabla \cdot (\Ub \rho h) + \frac{\partial p_0}{\partial t} + (\Ub \cdot \eb_r) \frac{\partial p_0}{\partial r}.
\end{equation}
For spherical geometry, we solve the analytically equivalent form,
\begin{equation}
\frac{\partial (\rho h)}{\partial t}  = - \nabla \cdot (\Ub \rho h) + \frac{\partial p_0}{\partial t} + \nabla \cdot (\Ub p_0) - p_0 \nabla \cdot \Ub.
\end{equation}
The update consists of two steps:

@@

\item Evolve $(\rho h)^{(1)} \rightarrow (\rho h)^{(2),\star}$.
\begin{description}
\item For planar geometry,
\begin{eqnarray}
(\rho h)^{(2),\star}
&=& (\rho h)^{(1)} \nonumber \\
&&- \dt \left\{ \nabla \cdot \left[ \uadvone (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \left(p_0^{n+1,*} - p_0^n\right) \nonumber \\
&& + \dt \left(\uadvone \cdot \eb_r\right) \left(\frac{\partial p_0}{\partial r} \right)^{n}
\end{eqnarray}

\item For spherical geometry,
\begin{eqnarray}
(\rho h)^{(2),\star}
&=& (\rho h)^{(1)} \nonumber \\
&&- \dt \left\{ \nabla \cdot \left[ \uadvone (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \left(p_0^{n+1,*} - p_0^n\right) \nonumber \\
&& + \dt \left \{ \nabla \cdot \left (\uadvone p_0^{n+\myhalf} \right ) - p_0^{n+\myhalf} \nabla \cdot \uadvone \right \}
\end{eqnarray}
\end{description}
where we define $p_0^\nph$ as $p_0^\nph = (p_0^n+p_0^{n+1,*})/2$.

\end{enumerate}
Note that outside the cutoff density region, we compute enthalpy using the EOS.

@@

The procedure to construct $\uadvtwodag$ is identical to the Godunov procedure
for computing $\uadvonedag$ in {\bf Step 3}, but uses
the updated value $S^{\nph}$ rather than $S^{\nph,\star}$.
We note that $\uadvtwo$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^{\nph} \uadvtwo\right) =
\beta_0^{\nph} \left[S^{\nph} - \frac{1}{\gammaonebar^{n+\myhalf}p_0^{n+\myhalf}}\left(\frac{\partial p_0}{\partial t}\right)^{n+\myhalf}\right],\label{eq:div2}
\end{equation}
where
\begin{equation}
\beta_0^{\nph} = \frac{ \beta_0^n +  \beta_0^{n+1,\star} }{2},
\qquad
\gammaonebar^{\nph} = \frac{ \gammaonebar^n +  \gammaonebar^{n+1,\star} }{2}.
\end{equation}

@@

\item Update the enthalpy.  This step is identical to {\bf Step 4D} except we use
  the updated values $\uadvtwo, \rho_0^{n+1}, (\rho h)_0^{n+1}$, and $p_0^{n+\myhalf}$
  rather than
  $\uadvone, \rho_0^{n+1,\star}, (\rho h)_0^{n+1,\star}$, and $p_0^n$.
  In particular:

@@

\item Evolve $(\rho h)^{(1)} \rightarrow (\rho h)^{(2)}$.
\begin{description}
\item For planar geometry,
\begin{eqnarray}
(\rho h)^{(2)}
&=& (\rho h)^{(1)} - \dt \left\{ \nabla \cdot \left[ \uadvtwo (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \left( p_0^{n+1} - p_0^n \right)\nonumber \\
&& + \dt \left(\uadvtwo \cdot \eb_r\right) \left(\frac{\partial p_0}{\partial r} \right)^\nph,
\end{eqnarray}

\item For spherical geometry,
\begin{eqnarray}
(\rho h)^{(2)}
&=& (\rho h)^{(1)} - \dt \left\{ \nabla \cdot \left[ \uadvtwo (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \left(p_0^{n+1} - p_0^n\right)\nonumber \\
&& + \dt \left[ \nabla \cdot \left (\uadvtwo p_0^{\nph} \right ) - p_0^{\nph} \nabla \cdot \uadvtwo \right],
\end{eqnarray}
\end{description}
where we now define $p_0^\nph$ as $p_0^\nph = (p_0^n+p_0^{n+1})/2$.

\end{enumerate}
Note that outside the cutoff density region, we compute enthalpy using the EOS.

@@

Finally, we use an approximate nodal projection to define $\Ub^{n+1}$
from $\Ub^{n+1,\dagger},$  such that $\Ub^{n+1}$ approximately
satisfies the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{n+\myhalf}\right].\label{eq:div3}
\end{equation}
As part of the projection we also define the new-time perturbational pressure,
$\pi^\nph.$  This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.  Details of the approximate projection are given in Paper III.

>>>>

In stratified low Mach number models, due to the extreme variation in density, the velocity can become very large in low density regions at the edge of the star.
Throughout Papers II-V, we have employed two techniques to help control these dynamics without significantly affecting the dynamics in the convective region.
First, we use a cutoff density technique, where we hold the density constant outside a specified radius (typically near where the density is $\sim$4 orders of magnitude smaller than the largest densities in the simulation.
Second, we employ a sponge technique where we artificially damp the velocities near and beyond the cutoff region.
For more details, refer to Paper V and the previous references cited within.

To initialize the simulation we use the same procedure described in Paper III.
At the beginning of each simulation, we define $(\Ub,\rho X_k,\rho h)$.
We set initial values for $\Ub, \rho X_k$, and $\rho h$ and perform a sequence of projections 
(to ensure the velocity field satisfies the divergence constraint) 
followed by a small number of steps of the temporal advancement scheme to iteratively 
find initial values for $\pi^{n-\myhalf}$ and $S^0$ and $S^1$ for use in the first time step.
The new temporal integration strategy in MAESTROeX contains the following steps:
\begin{description}

@@

\item Update the enthalpy using a discretized version of equation
%
\begin{equation}
\frac{\partial(\rho h)}{\partial t} = -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc,
\end{equation}
%
again omitting the reaction terms since we already accounted for
them in {\bf React State}.  This equation takes the form:
\begin{equation}
\frac{\partial (\rho h)}{\partial t}  = - \nabla \cdot (\Ub \rho h) + \frac{\partial p_0}{\partial t} + (\Ub \cdot \eb_r) \frac{\partial p_0}{\partial r}.
\end{equation}
For spherical geometry, we solve the analytically equivalent form,
\begin{equation}
\frac{\partial (\rho h)}{\partial t}  = - \nabla \cdot (\Ub \rho h) + \frac{\partial p_0}{\partial t} + \nabla \cdot (\Ub p_0) - p_0 \nabla \cdot \Ub.
\end{equation}
The update consists of two steps:

@@

\item Evolve $(\rho h)^{(1)} \rightarrow (\rho h)^{(2),\star}$ using
\begin{equation}
(\rho h)^{(2),\star}
= (\rho h)^{(1)} - \dt \left\{ \nabla \cdot \left[ \uadvone (\rho h)^{\nph,\pred} \right] \right\} + \Delta t\frac{Dp_0}{Dt}
\end{equation}
where here
\begin{equation}
\frac{Dp_0}{Dt} =
\begin{cases}
\frac{p_0^{n+1,*} - p_0^n}{\Delta t} + \left(\uadvone \cdot \eb_r\right) \left(\frac{\partial p_0}{\partial r} \right)^{n}& {\rm (planar)} \\
\frac{p_0^{n+1,*} - p_0^n}{\Delta t} + \left[ \nabla \cdot \left (\uadvone p_0^{\nph} \right ) - p_0^{\nph} \nabla \cdot \uadvone \right]& {\rm (spherical)}
\end{cases}
,
\end{equation}
and $p_0^\nph = (p_0^n+p_0^{n+1,*})/2$.
\end{enumerate}
\end{enumerate}

@@

The procedure to construct $\uadvtwodag$ is identical to the Godunov procedure
for computing $\uadvonedag$ in {\bf Step 3}, but uses
the updated value $S^{\nph}$ rather than $S^{\nph,\star}$.
We note that $\uadvtwo$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^{\nph} \uadvtwo\right) =
\beta_0^{\nph} \left[S^{\nph} - \frac{1}{\gammaonebar^{\nph}p_0^{\nph}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right],\label{eq:div2}
\end{equation}
where
\begin{equation}
\beta_0^{\nph} = \frac{ \beta_0^n +  \beta_0^{n+1,\star} }{2},
\qquad
\gammaonebar^{\nph} = \frac{ \gammaonebar^n +  \gammaonebar^{n+1,\star} }{2}.
\end{equation}

@@

\item Update the enthalpy.  This step is identical to {\bf Step 4D} except we use
  the updated values $\uadvtwo, \rho_0^{n+1}, (\rho h)_0^{n+1}$, and $p_0^{\nph}$
  rather than
  $\uadvone, \rho_0^{n+1,\star}, (\rho h)_0^{n+1,\star}$, and $p_0^n$.
  In particular:

@@

\item Evolve $(\rho h)^{(1)} \rightarrow (\rho h)^{(2)}$.
\begin{equation}
(\rho h)^{(2)}
= (\rho h)^{(1)} - \dt \left\{ \nabla \cdot \left[ \uadvtwo (\rho h)^{\nph,\pred} \right] \right\} + \Delta t\frac{Dp_0}{Dt}
\end{equation}
where here
\begin{equation}
\frac{Dp_0}{Dt} =
\begin{cases}
\frac{p_0^{n+1} - p_0^n}{\Delta t} + \left(\uadvtwo \cdot \eb_r\right) \left(\frac{\partial p_0}{\partial r} \right)^{n}& {\rm (planar)} \\
\frac{p_0^{n+1} - p_0^n}{\Delta t} + \left[ \nabla \cdot \left (\uadvtwo p_0^{\nph} \right ) - p_0^{\nph} \nabla \cdot \uadvtwo \right]& {\rm (spherical)}
\end{cases}
,
\end{equation}
and $p_0^\nph = (p_0^n+p_0^{n+1})/2$.
\end{enumerate}
\end{enumerate}

@@

Finally, we use an approximate nodal projection to define $\Ub^{n+1}$
from $\Ub^{n+1,\dagger},$  such that $\Ub^{n+1}$ approximately
satisfies the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right].\label{eq:div3}
\end{equation}
As part of the projection we also define the new-time perturbational pressure,
$\pi^\nph.$  This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.  Details of the approximate projection are given in Paper III.
